Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-01 ms.tex
brak3

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the accretion rate as a function of black hole mass for the average star formation heating, calculated from equation (\ref{eq:eddr_analytic}) using our results for the total wind heating (Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim 10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.  These fall below the maximum thermally stable accretion rate $\dot{M}_{\rm TI}$ (yellow lines), except at the highest black hole masses $\sim 10^{9}M_{\odot}$, for which the older stellar population supplies insufficient stellar wind heating and SN Ia are too rare near the stagnation radius.  In fact, for these highest mass black holes, the maximum accretion rate set by SN Ia blow-out (green lines; eq.~[\ref{eq:eddr_Ia}]) could cap the accretion rate below the thermally-unstable value, provided that the Ia heating efficiency is sufficiently high that $v_{\rm w}^{\rm Ia} \gg \sqrt{2}\sigma \sim 400$ km s$^{-1}$ (see discussion just preceeding eq.~[\ref{eq:eddr_Ia}]).  This is only possible if the Ia heating efficiency is reasonably high $\epsilon_{\rm Ia} > 0.1$ than normally assumed (\citealt{Sharma+14}).  For lower mass black holes, the maximum Ia accretion rate is significantly higher and cannot prevent thermally unstable flows, i.e. $\dot{M}_{\rm Ia} > \dot{M}_{\rm TI}$.

Finally, also shown for comparison in Figure \ref{fig:bh_growth} (top panel) is the accretion rate onto SgrA* estimated by \citet{Quataert:2004a} and the upper limit onto NGC3115 set by \citealt{ShcherbakovWong+:2014a}.

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Critical values of the black hole accretion rate $\dot{M}$ in Eddington units as a function of black hole mass $M_{\bullet}$, shown separately for core ($\Gamma=0.1$, solid lines) and cusp galaxies ($\Gamma=0.8$, dot-dashed lines).  Black lines show the accretion rate calculated from equation (\ref{eq:eddr_analytic}) using the stellar wind calculated for the average star formation histories for each galaxy mass (Fig.~\ref{fig:vwSources}).  Green lines show the maximum accretion set by Ia supernovae blow-out (\ref{eq:eddr_Ia}).  Red lines show the Compton-regulated accretion rate (eq.~[\ref{eq:MdotC}]).  Yellow lines show the maximum accretion rate for a thermally stable flow near the stagnation radius (eq.~[\ref{eq:Mdotmax}]).  {\bf BDM: add SgrA* numbers derived by quataert, NGC3115 from Roman} {\it Bottom panel:} Growth times $\Mbh/\Mdot$ in units
  of the Hubble time $\th$ for each of the accretion rates shown in the top panel.}
\end{figure}

@@

The X-ray luminosities of galactic nuclei provide a powerful diagnostic of the SMBH accretion rates and how they vary across different galaxy properties and SMBH masses (e.g.,  \citealt{Ho08}  and references therein).  Figure~\ref{fig:miller} shows our predictions for $\langle L_{X} \rangle =\epsilon_X \epsilon_{\rm bol} \dot{M} c^2$ in Eddington units as a function of black hole mass, where $\epsilon_{\rm bol} = L_{X}/L_{\rm bol}= 0.1$ is the assumed bolometric correction appropriate for low luminosity AGN (\citealt{Ho08}).  We have averaged the accretion rate used in calculating $\langle L_X \rangle$ between the values for galaxies with core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) stellar density profiles, by fiting the average value of the inner density slope $\Gamma$ as a function of black hole mass from the \citet{LauerFaber+:2007a} galaxy sample.

The left panel of Figure \ref{fig:miller} is calculated assuming a constant low value for the radiative efficiency $\epsilon_X = 10^{-4}$, typical of those estimated for low luminosity AGN (e.g., \citealt{Ho:2009a}) and SgrA* (e.g., \citealt{Quataert+99}).  The luminosity shown in the right panel instead assumes an efficiency of $\epsilon_X= \dot{M}/\dot{M}_{\rm edd}$ that depends on $\dot{M}$ in the way predicted by models for radiatively inefficient accretion flows (\citealt{Narayan&Yi95}).  Plotted next to the theoretical curves are the X-ray measurements (green stars) and upper limits (gray triangles) compiled by \citet{Miller+15} (cf.~\citealt{Gallo+10}). A green line shows the resulting best power-law fit to the average X-ray luminosity, which is given by $\langle L_X/L_{\rm edd}\rangle \sim \Mbh^\alpha$ from \citet{Miller+15}, with $\alpha = XXX$  (see also \citep{Zhang+09, Pellegrini10, Gallo+10} for similar conclusions).  Yellow, teal, and red lines show the maximum accretion rates, respectively, for thermally stable accretion (eq.~[\ref{eq:Mdotmax}]), as set by SN Ia blow-out (eq.~[\ref{eq:eddr_Ia}]), and as allowed by Compton heating feedback (eq.~[\ref{eq:MdotC}]).

@@

If heating in the nuclei of elliptical galaxies is characterized by that of the average star formation history of galaxies with similar mass, then to first order we predict that $\langle L_{X}/L_{\rm edd}\rangle$ should increase with black hole mass, though only weakly for $M_{\bullet} \lesssim 3\times 10^{8}M_{\odot}$.  This prediction is in tension with the observed trend that average Eddington ratio decreases (also weakly) with $M_{\bullet}$.  However, the physics setting $L_X$ based on the large-scale accretion rate is notoriously uncertain and there could exist black hole mass-dependent efficiency factors we are not taking into account.  For instance, if disk disk outflows carry away a substantial fraction of the infalling mass before it reaches sufficiently small radii to produce nuclear X-rays (e.g.~\citealt{Blandford&Begelman99}), then the angular momentum of the infalling gas (and hence that of stars in the galactic nucleus) might also affect the X-ray luminosity, potentially in a way that would depend systematically on the black hole mass.

It is also important to ask whether the steady-state accretion picture can explain the large large scatter (typically 2-3 orders of magnitude) observed in $L_{X}/L_{\rm edd}$.  Large scatter can in fact result from two effects: (1) the order of magnitude difference between $\dot{M}$ for core and cusp galaxies at fixed $v_w$ and $M_{\bullet}$ results in scatter due to the scatter in the Nuker stellar density profile $\Gamma$; (2) the strong sensitivity of the accretion rate on the gas heating rate, $\dot{M}/\dot{M}_{\rm edd} \propto v_{w}^{-3.8(-2.4)}$ for core(cusp) galaxies, respectively (Fig.~\ref{fig:mdot_mass}; eq.~[\ref{eq:eddr_analytic}]) combined with the significant dependence of $v_w$ on stochastic intermittency in the star formation history (Fig.~NICK).  The combination of (1) and (2) could well explain much of the scatter observed.

However, it is also possible that many of the galactic nuclei from the \citet{Miller+14} sample are in a thermally stable state.  The efficiency show in the right panel of Figure \ref{fig:miller} in some ways represents the maximum efficiency, since it does not take into account outflows from the accretion flow that would suppress the accreiton rate reaching large radii.  Nevertheless, many of the observed galaxies at low black hole masses of $M_{\bullet} \lesssim 10^{7}M_{\odot}$ lie well below the predicted average value for stable accretion (black line) and even approaching or above the maximally stable accretion rate $\dot{M}_{\rm TI}$ (yellow line) .  This suggests that these galaxies experience lower levels of heating than average galaxies of their mass, or that they are not accreting stably at all ($\S\ref{sec:stable}$).

>>>>

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the accretion rate as a function of black hole mass for the average star formation heating, calculated from equation (\ref{eq:eddr_analytic}) using our results for the total wind heating (Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim 10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.  These fall below the maximum thermally stable accretion rate, $\dot{M}_{\rm TI}$ (plotted as yellow lines), except at the highest black hole masses $\sim 10^{9}M_{\odot}$.  For these highest mass black holes, the maximum accretion rate as set by gas blow-out from SN Ia (green lines; eq.~[\ref{eq:eddr_Ia}]) could instead cap the accretion rate below the thermally-unstable value, provided that the SN Ia heating efficiency is sufficiently high that $v_{\rm w}^{\rm Ia} \gg \sqrt{2}\sigma \sim 400$ km s$^{-1}$ (see discussion just preceeding eq.~[\ref{eq:eddr_Ia}]).  This is only possible if the Ia heating efficiency is higher $\epsilon_{\rm Ia} > 0.1$ than normally assumed (\citealt{Sharma+14}).  For lower mass black holes, the maximum accretion rate set by SN Ia is significantly higher and cannot prevent the development of thermally unstable flows, i.e. $\dot{M}_{\rm Ia} \gg \dot{M}_{\rm TI}$.

Shown also for comparison in Figure \ref{fig:bh_growth} (top panel) is the accretion rate onto SgrA* estimated by \citet{Quataert:2004a} and the upper limit on $\dot{M}$ for the black hole in NGC3115 set by \citealt{ShcherbakovWong+:2014a}.  These points are consistent with our thermally-stable accretion rates within the expected dispersion (see next section), although NGC3115 is in fact a low luminosity AGN.

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Critical values of the black hole accretion rate $\dot{M}$ in Eddington units as a function of black hole mass $M_{\bullet}$, shown separately for core ($\Gamma=0.1$, solid lines) and cusp galaxies ($\Gamma=0.8$, dot-dashed lines).  Black lines show the accretion rate calculated from equation (\ref{eq:eddr_analytic}) using our estimate of the gas heating rate provided at $z = 0$ by the average star formation histories for each galaxy mass (Fig.~\ref{fig:vwSources}).  Green lines show the maximum accretion set by Ia supernovae blow-out (\ref{eq:eddr_Ia}).  Red lines show the accretion rate so obtained if Compton heating alone was acting (eq.~[\ref{eq:MdotC}]).  Yellow lines show the maximum accretion rate for a thermally stable flow near the stagnation radius (eq.~[\ref{eq:Mdotmax}]).  {\bf BDM: add SgrA* numbers derived by quataert, NGC3115 from Roman} {\it Bottom panel:} Growth times $\Mbh/\Mdot$ in units
  of the Hubble time $\th$ for each of the accretion rates shown in the top panel.}
\end{figure}

@@

The unresolved core X-ray luminosities of galactic nuclei provide a powerful diagnostic of the SMBH accretion rates and how they vary across different galaxy properties and SMBH masses (e.g.,  \citealt{Ho08}  and references therein).  Figure~\ref{fig:miller} shows our predictions for $\langle L_{X} \rangle =\epsilon_X \epsilon_{\rm bol} \dot{M} c^2$ in Eddington units as a function of black hole mass, where $\epsilon_{\rm bol} = 0.1$ is the approximate bolometric correction at X-ray energies appropriate for low luminosity AGN (\citealt{Ho08}).  We have averaged the accretion rate used in calculating $\langle L_X \rangle$ between the values for galaxies with core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) stellar density profiles using the average value of the inner density slope $\Gamma$ as a function of black hole mass as fit from the \citet{LauerFaber+:2007a} galaxy sample, an exercise that results in the relationship $\Gamma = ...$. {\bf BDM: fill in}

The left panel of Figure \ref{fig:miller} is calculated assuming a constant low value for the radiative efficiency $\epsilon_X = 10^{-4}$ typical of those estimated for low luminosity AGN (e.g., \citealt{Ho:2009a}), while the right panel luminosities are calculated instead assuming an efficiency of $\epsilon_X= \dot{M}/\dot{M}_{\rm edd}$ that depends on the Eddington ratio as predicted by some models for radiatively inefficient accretion flows (\citealt{Narayan&Yi95}; \citealt{Narayan+98}).  Shown for comparison are the X-ray measurements (green stars) and upper limits (gray triangles) of early-type galaxies compiled by \citet{Miller+15} (cf.~\citealt{Gallo+10}).  A green line shows the best power-law fit to the X-ray luminosity from \citet{Miller+15}, given by $\langle L_X/L_{\rm edd}\rangle \sim \Mbh^\alpha$ with $\alpha = XXX$  (see also \citep{Zhang+09, Pellegrini10, Gallo+10} for similar conclusions).  Yellow, teal, and red lines show the maximum accretion rates, respectively, for thermally stable accretion (eq.~[\ref{eq:Mdotmax}]), as set by SN Ia blow-out (eq.~[\ref{eq:eddr_Ia}]), and as allowed by Compton heating feedback (eq.~[\ref{eq:MdotC}]).

@@

If heating in the nuclei of elliptical galaxies is similar to that produced by the average star formation history of galaxies with similar mass, then to first order we predict that $\langle L_{X}/L_{\rm edd}\rangle$ should increase with black hole mass, though only weakly for $M_{\bullet} \lesssim 3\times 10^{8}M_{\odot}$.  This result is in tension with the observed down-sizing trend that the average Eddington ratio is observed to decrease (albeit also weakly) with $M_{\bullet}$.  However, in making such comparisons it is important to keep in mind the enormous uncertainty in calculating the luminosity of the accretion flow close to the black hole in a single waveband based on the accretion rate on larger scales.  Black hole mass-dependent efficiency factors could enter that we are not taking into account.  For instance, if disk outflows carry away a substantial fraction of the infalling mass before it reaches sufficiently small radii to produce the observed X-rays (e.g.~\citealt{Blandford&Begelman99}), then the angular momentum of the infalling gas (and hence that of stars in the galactic nucleus) might also affect the X-ray luminosity, potentially in a way that could depend systematically on the black hole mass.

Perhaps a more readily addressable test is whether a steady-state accretion picture can explain the large large scatter of typically 2-3 orders of magnitude observed in $L_{X}/L_{\rm edd}$.  Scatter can result from at least two effects (1) the order of magnitude difference between $\dot{M}$ for core and cusp galaxies at fixed $v_w$ and $M_{\bullet}$ (Fig.~\ref{fig:bh_growth}) will result in a similar scatter of $\dot{M}$ due to the variation in the Nuker stellar density profile $\Gamma$ at fixed $M_{\bullet}$; (2) the strong sensitivity of the accretion rate to the stellar wind velocity, $\dot{M}/\dot{M}_{\rm edd} \propto v_{w}^{-3.8(-2.4)}$ for core(cusp) galaxies, respectively (Fig.~\ref{fig:mdot_mass}; eq.~[\ref{eq:eddr_analytic}]) in combination with the significant dependence of $v_w$ on stochastic intermittency in the star formation history (Fig.~NICK).  The combination of (1) and (2) could well explain much of the scatter observed.

However, it is also possible that many of the galactic nuclei from the \citet{Miller+15} sample are in a thermally stable state.  The efficiency show in the right panel of Figure \ref{fig:miller} in some ways represents the maximum efficiency, since it does not take into account outflows from the accretion flow that would suppress the accreiton rate reaching large radii.  Nevertheless, many of the observed galaxies at low black hole masses of $M_{\bullet} \lesssim 10^{7}M_{\odot}$ lie well below the predicted average value for stable accretion (black line) and even approaching or above the maximally stable accretion rate $\dot{M}_{\rm TI}$ (yellow line) .  This suggests that these galaxies experience lower levels of heating than average galaxies of their mass, or that they are not accreting stably at all ($\S\ref{sec:stable}$).
